---
title: "Using Hind/He to identify paralogs"
author: "Lindsay Clark"
date: "September 4, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset and code for this example

We'll load the example *Miscanthus sinensis* diversity panel data that comes
with polyRAD.

```{r message = FALSE}
library(polyRAD)
myfile <- system.file("extdata", "Msi01genes.vcf", package = "polyRAD")
myRAD <- myRAD <- VCF2RADdata(myfile, possiblePloidies = list(2))
```

We'll also load the R function for estimating Hind/He.  This is a wrapper for
an Rcpp function implemented in `simpson.cpp`.  I'll plan to add this to polyRAD
once I have a preprint up for the methodology.

```{r}
source("hindhe.R")
```

## Estimating Hind/He

We will now run the function on the dataset.  By default it will ignore any
blank taxa.

```{r}
myHindHe <- HindHe(myRAD)
myHindHeByLoc <- colMeans(myHindHe, na.rm = TRUE)
```

## Interpreting the results

We can look at the distribution of values.

```{r}
hist(myHindHeByLoc)
abline(v = 0.5, col = "red")
```

With no inbreeding or population structure, the expected value is 
(ploidy - 1)/ploidy, which in this case is 0.5 for a diploid.
(The expected value for an autotetraploid is 0.75).  Inbreeding and
population structure lower this expected value slightly.  With a larger
set of markers, you can expect to see more well defined peaks, giving you a
better estimate of what value is reasonable and what markers should be
filtered out.

Let's look at individual markers.

```{r}
myHindHe
```

S01_150928 has a value close to what was expected.  Let's look at read depth
distribution for that marker.

```{r}
myRAD$alleleDepth[1:50,myRAD$alleles2loc == match("S01_150928", GetLoci(myRAD))]
```

There's a little bit of contamination, but generally when we see an apparently
heterozygous individual, the read depth ratios are fairly even and only belong
to two alleles.

S01_535898 has a value much higher than expected, and S01_535839 much lower than
expected.  In both of these cases, the read depth may simply be too low to get
an accurate estimate.

```{r}
myRAD$alleleDepth[1:50,myRAD$alleles2loc == match("S01_535898", GetLoci(myRAD))]
```

```{r}
myRAD$alleleDepth[1:50,myRAD$alleles2loc == match("S01_535839", GetLoci(myRAD))]
```

## A simulated example

To artificially create some paralogous loci, let's scramble the assignment of
alleles to loci.

```{r}
set.seed(904) # to keep the same randomization.
myRAD2 <- myRAD
myRAD2$alleles2loc <- sample(myRAD2$alleles2loc)
myHindHe2 <- HindHe(myRAD2)
myHindHeByLoc2 <- colMeans(myHindHe2, na.rm = TRUE)
hist(myHindHeByLoc2, breaks = 10)
```

Now let's look at the marker with the highest value.

```{r}
highmarker <- which(myHindHeByLoc2 > 0.8)
hmdepth <- myRAD2$alleleDepth[,myRAD2$alleles2loc == highmarker]
hmdepth[rowSums(hmdepth) > 0,][1:50,]
```

Here we see an overabundance of apparent heterozygotes, and they don't follow
expected read depth ratios very well.
